---
title: "New York Times Headlines"
format: 
  html:
    code-fold: true
    
execute:
  message: false
  warning: false
  echo: true
---

Data source: *New York Times Annotated Corpus*, distributed with the **RTextTools** package

Timothy P. Jurka, Loren Collingwood, Amber E. Boydstun, Emiliano Grossman and Wouter van Atteveldt (2012). RTextTools: Automatic Text Classification via Supervised Learning. R package version 1.3.9. http://CRAN.R-project.org/package=RTextTools.\
Accessed via built-in dataset `NYTimes` in R.

```{r}
library(tidyverse)
library(stringr)
library(lubridate)
library(RTextTools) 
library(forcats)
data(NYTimes)
nyt<-as_tibble(NYTimes)

nyt<-nyt |>
  mutate(Date=as.Date(Date, format = "%d-%b-%y"))

nyt_clean <- nyt |>
  mutate(clean_headline=Title |>
           str_to_lower()|>
           str_replace_all("[^a-z\\s]", " ")|>
           str_squish(),
         year=year(Date)
        )

nyt_mentions <- nyt_clean |>
  mutate(
    clinton = str_detect(clean_headline, "(?<![a-z])clinton(?![a-z])"),
    bush = str_detect(clean_headline, "(?<![a-z])bush(?![a-z])")
  )

nyt_periods <- nyt_mentions |>
  mutate(
    period = case_when(
      Date >= as.Date("1997-01-20") & Date < as.Date("2001-01-20") ~ "Clinton Era 
      (01/20/1997 to 01/20/2001)",
      Date >= as.Date("2001-01-20") & Date < as.Date("2005-01-20") ~ "Bush era 
      (01/20/2001 to 01/20/2005)",
      TRUE ~ NA_character_
    )
  ) |>
  filter(!is.na(period))

nyt_counts <- nyt_periods |>
  group_by(period) |>
  summarise(
    clinton_mentions = sum(clinton),
    bush_mentions = sum(bush)
  )

  

nyt_plot<-nyt_counts|>
  pivot_longer(cols = c(clinton_mentions, bush_mentions),
               names_to = "name", values_to = "count") 
 

ggplot(nyt_plot, aes(x = period, y = count, fill = name)) +
  geom_col(position = "dodge") +
  labs(
    title = "NYT Headline Mentions: Clinton vs Bush",
    x = "Presidential Era", 
    y = "Number of Mentions",
    fill = "President"
  ) +
  scale_fill_manual(
    values = c("clinton_mentions" = "blue", "bush_mentions" = "red"),
    labels = c("clinton_mentions" = "Clinton mentions",
               "bush_mentions" = "Bush mentions")
  ) +
  theme_minimal()

```

This bar plot compares the total number of NYT headlines mentioning Clinton and Bush during their presidential terms. It shows that Bush received more media coverage overall than Clinton. This visualization highlights the relative attention each president received from the NYT and allows for a clear comparison between the two eras.

```{r}
nyt_trend <- nyt_mentions |>
  filter(Date >= as.Date("1997-01-20") & Date < as.Date("2005-01-20")) |>
  mutate(month = floor_date(Date, "month")) |>
  group_by(month) |>
  summarise(
    clinton = sum(clinton),
    bush    = sum(bush)
  ) |>
  pivot_longer(cols = c(clinton, bush),
               names_to = "name",
               values_to = "count")

ggplot(nyt_trend, aes(x = month, y = count, color = name)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_x_date(date_labels = "%b %Y", date_breaks = "6 months") + 
  scale_color_manual(
    values = c("clinton" = "blue", "bush" = "red"),
    labels = c("clinton" = "Clinton mentions",
               "bush" = "Bush mentions")) +
  
  labs(
    title = "Monthly NYT Headline Mentions: Clinton vs Bush",
    x = "Month",
    y = "Number of Mentions",
    color = "President"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size=8)) 
  


```

This line plot shows how mentions of Clinton and Bush changed month by month throughout their presidencies. Spikes in the lines correspond to major events or news stories, such as elections, inaugurations, or controversies. It provides insight into the timing and dynamics of media attention, revealing trends that are not visible in the total counts alone.
